---
title: "Homework 2"
author: "Jiaxuan Su"
output: github_document
---

```{r setup, message=FALSE}
library(tidyverse)
```

## Problem 1

```{r, message=FALSE}
subway =
  read_csv(
    "/Users/sjxx/Downloads/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    col_types = cols(
      Route8 = col_character(),
      Route9 = col_character(),
      Route10 = col_character(),
      Route11 = col_character()
    )
  ) |>
  janitor::clean_names() |>
  select(line:route11, entry, vending, entrance_type, ada) |>
  mutate(entry = entry == "YES")
```

The data include station names, lines, coordinates, routes, entry, vending, entrance type, and ADA status. I cleaned the names, kept the needed columns, read routes as character, and changed entry to logical. There are `r nrow(subway)` rows and `r ncol(subway)` columns. The data are not tidy because routes are in separate columns.

```{r}
# distinct stations
subway |>
  distinct(line, station_name) |>
  nrow()

# ADA compliant stations
subway |>
  filter(ada) |>
  distinct(line, station_name) |>
  nrow()

# proportion allowing entry without vending
subway |>
  filter(vending == "NO") |>
  pull(entry) |>
  mean()
```

```{r}
subway_long =
  subway |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name"
  ) |>
  mutate(route_number = as.numeric(route_number)) |>
  drop_na(route_name)

# stations serving A train
subway_long |>
  filter(route_name == "A") |>
  distinct(line, station_name) |>
  nrow()

# ADA compliant stations serving A train
subway_long |>
  filter(route_name == "A", ada) |>
  distinct(line, station_name) |>
  nrow()
```

## Problem 2

```{r}
trash_file = "/Users/sjxx/Downloads/202610 Trash Wheel Collection Data.xlsx"

mr_trash =
  readxl::read_excel(
    trash_file, sheet = "Mr. Trash Wheel", range = "A2:N743"
  ) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    year = as.numeric(year),
    sports_balls = as.integer(round(sports_balls)),
    wheel = "Mr. Trash Wheel"
  )

professor =
  readxl::read_excel(
    trash_file, sheet = "Professor Trash Wheel",
    range = "A2:M140", na = c("", "dive")
  ) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(wheel = "Professor Trash Wheel")

gwynnda =
  readxl::read_excel(
    trash_file, sheet = "Gwynnda the Good Wheel of the W",
    range = "A2:L395"
  ) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(wheel = "Gwynnda")

trash = bind_rows(mr_trash, professor, gwynnda)

professor_weight = sum(professor$weight_tons, na.rm = TRUE)

gwynnda_butts =
  gwynnda |>
  filter(month == "June", year == 2022) |>
  pull(cigarette_butts) |>
  sum(na.rm = TRUE)

nrow(trash)
professor_weight
gwynnda_butts
```

For problem 2, I combined the three wheels into `r nrow(trash)` observations. Each row is one dumpster collection, with variables such as wheel, date, weight, and cigarette butts. Professor Trash Wheel collected `r round(professor_weight, 2)` tons of trash. Gwynnda collected `r format(gwynnda_butts, scientific = FALSE)` cigarette butts in June 2022.

## Problem 3

```{r, message=FALSE}
baseline_raw =
  read_csv("/Users/sjxx/Downloads/data_mci/MCI_baseline.csv", skip = 1, na = ".") |>
  janitor::clean_names()

baseline =
  baseline_raw |>
  filter(is.na(age_at_onset) | age_at_onset > current_age) |>
  mutate(
    sex = case_match(sex, 0 ~ "Female", 1 ~ "Male"),
    apoe4 = case_match(apoe4, 0 ~ "Non-carrier", 1 ~ "Carrier")
  )

women = filter(baseline, sex == "Female")
```

I skipped the notes row, treated dots as missing, and labeled sex and APOE4. After excluding baseline MCI cases, I kept `r nrow(baseline)` of `r nrow(baseline_raw)` recruits. Of these, `r sum(!is.na(baseline$age_at_onset))` developed MCI. Mean baseline age was `r round(mean(baseline$current_age), 2)` years, and `r mean(women$apoe4 == "Carrier") * 100`% of women carried APOE4.

```{r, message=FALSE}
amyloid =
  read_csv("/Users/sjxx/Downloads/data_mci/mci_amyloid.csv", skip = 1) |>
  janitor::clean_names() |>
  rename(id = study_id, time_0 = baseline) |>
  pivot_longer(
    time_0:time_8,
    names_to = "time",
    names_prefix = "time_",
    values_to = "amyloid_ratio"
  ) |>
  mutate(time = as.numeric(time))
```

I skipped the notes row and put each visit in a separate row. The data have `r nrow(distinct(amyloid, id))` people and `r nrow(amyloid)` rows, with ID, visit time, and amyloid ratio. Visits cover 0–8 years, and missing ratios are kept.

```{r}
baseline_only = filter(baseline, !(id %in% amyloid$id))

amyloid_only =
  amyloid |>
  filter(!(id %in% baseline$id)) |>
  distinct(id)

mci = inner_join(baseline, amyloid, by = "id")

write_csv(mci, "/Users/sjxx/Downloads/data_mci/mci_combined.csv")
```

I found `r nrow(baseline_only)` people only in the cleaned baseline data and `r nrow(amyloid_only)` only in amyloid. I joined matching IDs, giving `r nrow(distinct(mci, id))` people, `r nrow(mci)` rows, and `r ncol(mci)` columns. Each row combines demographics and biomarker data for one visit.